---
title: "Practical Machine Learning Project"
author: "Oliver Ledesma"
date: "Saturday, December 20, 2014"
output: html_document
---

Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here

Data

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.
The training data are available here and the test data here

We download the two datasets

```{r, echo=FALSE}
# check if a data folder exists; if not then create one
if (!file.exists("data/pml-training.csv")) {dir.create("data")

# file URL and destination file
fileUrl1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
destfile1 <- "./data/pml-training.csv"
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
destfile2 <- "./data/pml-testing.csv"


# download the file and note the time
download.file(fileUrl1, destfile = destfile1)
download.file(fileUrl2, destfile = destfile2)
dateDownloaded <- date()
}

train <- read.csv("./data/pml-training.csv", na.strings= c("NA",""," "))
test <- read.csv("./data/pml-testing.csv", na.strings= c("NA",""," "))
```

```{r}
dim(train)
table(train$classe)
```

Preprocessing

Partitioning the training set

We separate our training data into a training set and a validation set so that we can validate our model.

```{r}
library(caret)
```

```{r}
set.seed(123456)
trainset <- createDataPartition(train$classe, p = 0.8, list = FALSE)
Training <- train[trainset, ]
Validation <- train[-trainset, ]
```

Feature Cleanup
Exclude near zero variance, columns with 40% or more missing values and descriptive columns.

```{r}
# exclude near zero variance features
nzvcol <- nearZeroVar(Training)
Training <- Training[, -nzvcol]

# exclude columns with 40% or more missing values exclude descriptive
# columns like name etc
cntlength <- sapply(Training, function(x) {
    sum(!(is.na(x) | x == ""))
})
nullcol <- names(cntlength[cntlength < 0.6 * length(Training$classe)])
descriptcol <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", 
    "cvtd_timestamp", "new_window", "num_window")
excludecols <- c(descriptcol, nullcol)
Training <- Training[, !names(Training) %in% excludecols]
```

Use random forest as model for classification and regression.

```{r}
library(randomForest)
rfModel <- randomForest(classe ~ ., data = Training, importance = TRUE, ntrees = 10)
```

Test Model

```{r}
ptraining <- predict(rfModel, Training)
print(confusionMatrix(ptraining, Training$classe))
```

Validation set accuracy (Out-of-Sample)

```{r}
pvalidation <- predict(rfModel, Validation)
print(confusionMatrix(pvalidation, Validation$classe))
```

Test set prediction

```{r}
ptest <- predict(rfModel, test)
ptest
```

Write answers to text files.

```{r}
answers <- as.vector(ptest)

pml_write_files = function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
            col.names = FALSE)
    }
}

pml_write_files(answers)
```